INGRID: Creating Languages in MPS from ANTLR Grammars

### 2016-10-10 chapters/background.tex
editor and textgen

\paragraph{Structure.}
The fundamental aspect of any MPS language is Structure.
It must be created first for each intended element (concept) of the language.
Structure specifies core attributes of an MPS concept such as the name, inheritance relationships, possible child concepts (including their types and cardinalities), the implemented interfaces, and references to other AST nodes.
% other properties (fields of any type that can hold values), etc.
Figure~\ref{fig:if_statement_structure} shows definition of the Structure aspect for the \code{if} statement.

@@

\begin{figure}[ht]
\centering
\begin{alltt}
\small
\mpskeyword{concept} IfStatement \mpskeyword{extends} Statement
        \mpskeyword{implements} IContainsStatementList
                   IDontSubstituteByDefault
                   IConditional

  \mpskeyword{instance can be root:} false
  \mpskeyword{alias:} \mpsalias{if}
  \mpskeyword{short description:} \mpsplaceholder{<no short description>}

  \mpskeyword{properties:}
  \mpsproperty{forceOneLine}   : boolean
  \mpsproperty{forceMultiLine} : boolean

  \mpskeyword{children:}
  \mpsproperty{condition}        : Expression[\mpscardinality{1}]
  \mpsproperty{ifFalseStatement} : Statement[\mpscardinality{0..1}]
  \mpsproperty{ifTrue}           : StatementList[\mpscardinality{1}]
  \mpsproperty{elsifClauses}     : ElsifClause[\mpscardinality{0..n}]

  \mpskeyword{references:}
  \mpsplaceholder{<< ... >>}
\end{alltt}
\caption{Structure aspect of the \code{if} statement}
\label{fig:if_statement_structure}
\end{figure}

@@

\paragraph{Editor.}
\JB{Odstavec mirne upraven}The editor aspect is where the user defines what the projectional editor representation of a code fragment (an AST node) looks like on the screen and how the user interacts with the code. JetBrains have developed a cellular system that allows placing node's (concept's) properties and children into different cells. The author usually incorporates all concept's children, references, and properties inside the representation so that the future users of the language can insert the values that the node expects. Additionally, all visual cells of the editor can be styled using a~language similar to CSS.

\begin{figure}[ht]
	\centering
	\includegraphics[scale=0.5]{./images/if_statement_editor_definition.png}
	\caption{"If statement" editor aspect definition}
	\label{fig:if_editor_definition}
\end{figure}

\JB{Nasledujici odstavec navrhuji odstranit}
\todo{From thesis.

The editor is a module dedicated/responsible for the definition of appearance of an AST node (concept).
JetBrains have developed a cellular system that allows placing node's (concept's) properties and children into different cells.
These cells then can be styled to user's liking.
There are many different types of cells, each behaving a little bit different towards its contents.
For example, MPS supports various horizontal and vertical lists, etc.
Extra cells can be added on top --- cells which just specify layout adjustments such as indentation and text color.
In Figure~\ref{fig:if_editor_definition}, you can see what a real editor definition might look like for the if statement of the Java language (as defined in MPS).
}

\paragraph{BaseLanguage}
\todo{Introduce the base language here.}
\JB{Odstavec prepsan}
In order to implement MPS itself and also to support the basic set of language-definition DSLs, BaseLanguage was developed in the early days of MPS. BaseLanguage is a clone of Java implemented using the MPS constructs. Although BaseLanguage is syntactically almost identical to Java, it is edited in a projectional editor, just like all languages in MPS. The DSLs that language authors use in order to define languages are generated into BaseLanguage. Similarly, all customer DSLs that are meant to be generated into Java choose BaseLanguage as their generation target and the conversion to textual Java will be handled by BaseLanguage without any further effort, since BaseLanguage has a TextGen aspect defined, which translates BaseLanguage code into textual Java sources.

\paragraph{TextGen.}
\JB{Odstavec upraven} The TextGen aspect defines how a certain AST node will be translated into text. The TextGen aspect is typically only needed for the bottom-line (base) languages. DSLs, on the other hand, need to define rules for model-to-model conversion (a generator), since these are rarely converted to text directly. After TextGen has generated textual sources from an AST, a compiler for the particular GPL gets invoke in order to compile the generate sources into binary code.
The TextGen definition follows a very straightforward pattern - each node outputs its textual representation into a textual buffer, calling TextGen of its children nodes at the right moments. MPS then calls this method for the root AST nodes of the given program.
TextGen aspect has to be defined (especially its more complex functionality) using the BaseLanguage.
Again, we included an example of the if statement (Figure~\ref{fig:if_statement_textgen}).

\begin{figure}[ht]
	\centering
	\includegraphics[scale=0.6]{./images/if_statement_textgen.png}
	\caption{"If statement" TextGen aspect definition}
	\label{fig:if_statement_textgen}
\end{figure}

\JB{Navrhuji odebrat}
\todo{From thesis.

After the program's AST is created using the projectional editor, there can be rules defined on how to generate code in plain text out of that tree.
This module is called "text gen".
These defined generators can target any language and platform.

\paragraph{Other aspects.}
MPS supports also other aspects, such as type checking, model transformations, static and data-flow code analysis, refactorings, language migrations, etc., but these are not important for our contribution. Moreover, the information needed for their automated generation is not available/contained inside a grammar file.

}

>>>>

\paragraph{Structure.}
The fundamental aspect of any MPS language is Structure.
It must be created first for each intended element (concept) of the language.
Structure specifies core attributes of an MPS concept such as the name, inheritance relationships, possible child concepts (including their types and cardinalities), the implemented interfaces, and references to other AST nodes.
% other properties (fields of any type that can hold values), etc.
Figure~\ref{fig:if_statement_structure} shows definition of the Structure aspect for the \code{if-else} statement.
\PP{Mozna trochu vic popsat obsah figure --- co znamenaji jednotlive prvky definice.}

@@

\begin{figure}[ht]
\centering
\begin{alltt}
\small
\mpsstkeyword{concept} IfStatement \mpsstkeyword{extends} Statement
        \mpsstkeyword{implements} IContainsStatementList
                   IDontSubstituteByDefault
                   IConditional

  \mpsstkeyword{instance can be root:} false
  \mpsstkeyword{alias:} \mpsstalias{if}
  \mpsstkeyword{short description:} \mpsstplaceholder{<no short description>}

  \mpsstkeyword{properties:}
  \mpsstproperty{forceOneLine}   : boolean
  \mpsstproperty{forceMultiLine} : boolean

  \mpsstkeyword{children:}
  \mpsstproperty{condition}        : Expression[\mpsstcardinality{1}]
  \mpsstproperty{ifFalseStatement} : Statement[\mpsstcardinality{0..1}]
  \mpsstproperty{ifTrue}           : StatementList[\mpsstcardinality{1}]
  \mpsstproperty{elsifClauses}     : ElsifClause[\mpsstcardinality{0..n}]

  \mpsstkeyword{references:}
  \mpsstplaceholder{<< ... >>}
\end{alltt}
\caption{Structure aspect of the \code{if-else} statement}
\label{fig:if_statement_structure}
\end{figure}

\paragraph{Editor.}
The Editor aspect is where the language designer specifies what the projectional editor representation of a code fragment (an AST) looks like on the screen and how the user interacts with the code.
JetBrains have developed a cellular system that enables placing properties and children of a node (concept) into different cells.
The author usually incorporates all of the node's children, references, and properties inside the representation, so that the future users of the language can insert all values that the node expects.
Additionally, cells of the editor can be styled using a language similar to CSS.
The supported visual characteristics include text color and indentation.

Figure~\ref{fig:if_editor_definition} shows an example of what the definition of the Editor aspect for the \code{if-else} statement might look like.
While here we indicate positions of cell borders on each line by spaces (just for the purpose of illustration), MPS GUI actually uses a graphically much more appealing way of displaying the Editor aspects, which involves vertical and horizontal lines of different colors and also background colors other than white for some cells.
\PP{Vysvetlit symboly "[-", "-]" \%, a dalsi ktere nejsou hned zrejme.}

\begin{figure}[ht]
\centering
\begin{alltt}
\small
\mpsedannotation{<default>} \mpsedkeyword{editor for concept} \mpsedtarget{IfStatement}
  \mpsedkeyword{node cell layout:}
    [-
      \mpsedkeyword{if} \textbf{(} \% conditions \% \textbf{)} [-
      \textbf{\{}
      [- \% ifTrue % -]
      \textbf{\}}
    -]
    ?[- \mpsedkeyword{else} \% ifFalseStatement \% -]
    -]
\end{alltt}
\caption{Editor aspect for the \code{if-else} statement}
\label{fig:if_editor_definition}
\end{figure}

\paragraph{BaseLanguage.}
Another important feature of MPS that we need to describe in more detail is \emph{BaseLanguage}.
It is a clone of Java implemented using the MPS constructs.
BaseLanguage was developed in the early days of MPS in order to implement MPS itself and also to support the basic set of language-definition DSLs,
Although BaseLanguage is syntactically almost identical to Java, it is edited in a projectional editor, just like all the languages in MPS.
The language-definition DSLs, used to define custom languages by their authors, are generated into the BaseLanguage.
Similarly, all the custom DSLs that are meant to be generated into Java choose BaseLanguage as their generation target.
The conversion to textual Java is handled by BaseLanguage without any further manual effort, since BaseLanguage has a TextGen aspect defined, which translates BaseLanguage code into textual Java sources.

\paragraph{TextGen.}
The TextGen aspect specifies how a given AST node will be translated into plain text representation.
It is typically needed only for the bottom-line base languages.
DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since these are rarely converted to text directly.
After TextGen has generated textual sources from an AST, a compiler for the particular GPL is invoked to compile the textual sources into binary code.
The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, calling TextGen of its children nodes at the right moments.
MPS calls the corresponding method on the root AST nodes of the given program.

TextGen aspect for each AST node (concept of the language) has to be defined using the BaseLanguage.
Again, we include an example for the \code{if-else} statement (Figure~\ref{fig:if_statement_textgen}).

\begin{figure}[ht]
\centering
\begin{alltt}
\small
\mpstgkeyword{text gen component for concept} \mpstgtarget{IfStatement} \{
  \mpstgparam{(context, buffer, node)->void} \{
    \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n;}}
    \mpstgaction{indent buffer;}
    \mpstgaction{append} \{\mpstgliteral{if (}\} \$\{\mpstgparam{node}.\mpstgnodeprop{condition}\} \{\mpstgliteral{) \{}\};
    \mpstgkeyword{with indent} \{
      \mpstgaction{append} \$\{\mpstgparam{node}.\mpstgnodeprop{ifTrue}\};
    \}
    \mpstgaction{append} \textcolor{Blue}{\textbf{\textbackslash{}n}} \{\mpstgliteral{\}}\} \$list\{\mpstgparam{node}.\mpstgnodeprop{elsifClauses}\};
    \mpstgkeyword{if} (\mpstgparam{node}.\mpstgnodeprop{ifFalseStatement}.\textbf{isNotNull}) \{
      \mpstgaction{append} \{ \mpstgliteral{else}\} \$\{\mpstgparam{node}.\mpstgnodeprop{isFalseStatement}\};
    \}
  \}
\}
\end{alltt}
\caption{TextGen aspect definition for the \code{if-else} statement}
\label{fig:if_statement_textgen}
\end{figure}

@@
